A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2025-02-24 paper/writeup.tex
minor edits

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets.
        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. Thus the maximal mapping for 4 should be to itself. 
        On the rest of the genome, all of nodes match with their identical counterpart.
        This makes the matched span $\match(T1,T2)=43$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
        Given the above matching, the inverse matching will match nodes $0$ through $7$, and node $9$ has no match 
        since its only possible match ($4 \in T1$) was not the best match from $T1 \to T2$. 
        This means the inverse matched span $\invmatch(T1,T2)=43$ and $\operatorname{TPR}(T1,T2)=\frac{43}{47}$.
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

@@

We used simulations of several scenarios.
To include the effects of heterogeneous recombination rate,
in some we used stdpopsim \citep{adrion2020stdpopsim} 
to simulate chromosome 1 of \textit{Canis familiaris}
using the CanFam3 genetic map from \citet{campbell2016}.
\emph{``Constant dog''} simulations simulated this chromosome in a 
population of (constant) size $10^4$.
\emph{``Expanding dog''} simulations were similar, but
used a discrete-time Wright--Fisher model to simulate
a small population of 100 that expanded to 1,000 individuals ten generations ago,
which then doubled every generation to reach 512,000 individuals.
All jobs for which runtime was recorded were executed on an Intel Xeon Gold 6148 processor.
Additionally, we compare accuracy between sequences modified from a ``true'' ARG
using our dissimilarity methods. 
These ARGs were simulated with an effective population size of 1,000
and recombination rate of $10^{-8}$ with between 10 and 1,000 samples
and a genome length between $10^6$ to $5\times 10^7$.

@@

The method to extend haplotypes described here is available through the \tskit{}
python and C APIs (\url{https://tskit.dev/tskit}) as \texttt{extend\_haplotypes};
methods to compare ARGs are implemented in the \tscompare{} python package
(\url{https://tskit.dev/tscompare}).
\comment{Scripts used to produce the results in this paper are available at
\url{https://github.com/XXX/TODO}.}

>>>>

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{matched span}, $\match(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets.
        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. Thus the maximal mapping for 4 should be to itself. 
        On the rest of the genome, all of nodes match with their identical counterpart.
        This makes the matched span $\match(T1,T2)=43$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
        Given the above matching, the inverse matching will match nodes $0$ through $7$, and node $9$ has no match 
        since its only possible match ($4 \in T1$) was not the best match from $T1 \to T2$. 
        This means the inverse matched span $\invmatch(T1,T2)=43$ and $\operatorname{TPR}(T1,T2)=\frac{43}{47}$.
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

@@

We used simulations of several scenarios.
To include the effects of heterogeneous recombination rate,
in some we used stdpopsim \citep{adrion2020stdpopsim} 
to simulate chromosome 1 of \textit{Canis familiaris}
using the CanFam3 genetic map from \citet{campbell2016}.
\emph{``Constant dog''} simulations simulated this chromosome in a 
population of (constant) size $10^4$.
\emph{``Expanding dog''} simulations were similar, but
used a discrete-time Wright--Fisher model to simulate
a small population of 100 that expanded to 1,000 individuals ten generations ago,
which then doubled every generation to reach 512,000 individuals.
All jobs for which runtime was recorded were executed on an Intel Xeon Gold 6148 processor.
Additionally, we compare accuracy between sequences modified from a ``true'' ARG
using our matched span methods. 
These ARGs were simulated with an effective population size of 10,000
and recombination rate of $10^{-8}$ with between 10 and 1,000 samples
and a genome length between $10^6$ to $5\times 10^7$.

@@

The method to extend haplotypes described here is available through the \tskit{}
python and C APIs (\url{https://tskit.dev/tskit}) as \texttt{extend\_haplotypes};
methods to compare ARGs are implemented in the \tscompare{} python package
(\url{https://tskit.dev/tscompare}).
Scripts used to produce the results in this paper are available at
\url{https://github.com/hfr1tz3/haplotypes-and-ancestral-recombination-graphs}.
